---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# plot2

<!-- badges: start -->
<!-- badges: end -->

A lightweight extension of the base R `plot` function, with support for
automatic grouping and legend handling, and several other enhancements.

## Installation

You can install the development version of **plot2** from
[GitHub](https://github.com/):

``` r
# install.packages("remotes")
remotes::install_github("grantmcdermott/plot2")
```

## Motivation

R users are spoiled for choice when it comes to visualization frameworks. The
options of course include **ggplot2** (in my opinion, the most important
graphics system of the last decade) and **lattice**, not to mention a 
bewildering array of extensions built around, on top, and in between these
amazing packages.

It is perhaps not surprising then that the base R graphics system sometimes gets
short shrift. This is unfortunate, because base R offers extraordinarily
powerful and flexible plotting facilities. Just type in `demo("graphics")` in
your R console to get an idea, or see these two excellent 
[introductory](https://github.com/karoliskoncevicius/tutorial_r_introduction/blob/main/baseplotting.md)
[tutorials](https://quizzical-engelbart-d15a44.netlify.app/2021-2022_m2-data-2_visu-2_practice#1). The downside of this power and flexibility is that plotting in base R
can require quite a lot of manual tinkering. A case in point is plotting grouped
data and then generating an appropriate legend. Doing this with the generic
`plot()` function can require several function calls (e.g., or a loop), fiddling
with your plot regions, and writing the legend out manually.

The goal of `plot2` is to remove this annoyance by providing a lightweight
extension of the base (2D) `plot` function, particularly as it applies to
scatter and line plots with grouped data. For example, `plot2` makes it easy to
plot different categories of a dataset in a single function call and highlight
these categories (groups) using modern colour palettes. Coincident with this
grouping support, `plot2` also produces automatic legends with scope for further
customization. While the package offers several other minor enhancements, it
tries as far as possible to be a drop-in replacement for the equivalent base
plot function. Users should be able to swap a valid `plot` call with `plot2`
without any changes to the expected output.

In summary, consider trying the **plot2** package if you are looking for a more
convenient version of the base R `plot` functionality. You can use pretty much
the same syntax and all of your theming elements should carry over too. It has
no other dependencies than base R itself, which might make it more suitable for
situations where dependency management is expensive (e.g., an R application 
running ina browser via [WebAssembly](https://docs.r-wasm.org/webr/latest/).)

## Examples

Let's load the package then walk through some examples.

```{r}
library(plot2)
```

As far as possible, `plot2` tries to be a drop-in replacement for regular `plot`
calls.

```{r}
op = par(mfrow = c(1, 2))

plot(0:10, main = "plot")
plot2(0:10, main = "plot2")
```

Similarly, we can plot elements from a data frame using either the atomic or
formula methods.

```{r}
op = par(mfrow = c(2, 2))

plot(airquality$Day, airquality$Temp, main = "plot")
plot(Temp ~ Day, data = airquality, main = "plot (formula)")
plot2(airquality$Day, airquality$Temp, main = "plot2")
plot2(Temp ~ Day, data = airquality, main = "plot2 (formula)")

par(op) # reset to a unique (single) plot frame
```

So far, so good. Where `plot2` starts to diverge from its base counterpart,
however, is when it comes to grouped data. In particular, `plot2` allows you to
characterize groups using either the `by` argument. Notice that we also get an
automatic legend (we'll return to that shortly).

```{r}
plot2(airquality$Day, airquality$Temp, by = airquality$Month)
```

Even more convenient is to use the equivalent formula syntax. Just place the
grouping variable after a vertical bar `|`.

```{r}
plot2(Temp ~ Day | Month, airquality)
```

At this point, experienced base plot users might protest that you _can_ colour
by groups using the `col` argument.^[E.g.,
`plot(airquality$Day, airquality$Temp, col = airquality$Month)`.] This is true,
but there are several points to push back on. First, you don't get an automatic
legend. Second, the base `plot` formula method doesn't specify the grouping
within the formula itself (not a deal-breaker, but particularly consistent in my
view). Third, and perhaps most importantly, this grouping doesn't carry over
to line plots (i.e., type="l"). Instead, you have to transpose your data and use
`mplot`.^[See
[this](https://stackoverflow.com/questions/10519873/how-to-create-a-line-plot-with-groups-in-base-r-without-loops)
old StackOverflow thread.]

We can contrast this with `plot2` where converting to a grouped line plot is a
simple argument change. I'll also use this as opportunity to highlight that
the legend responds accordingly to these plot element changes and can be
further customized. Note the use of the "!" in the `legend.position` argument
to place the legend _outside_ the plot area.

```{r}
plot2( 
  Temp ~ Day | Month, airquality,
  type = "l",
  legend.position = "right!"
)
```

Let's quickly pivot to colour palettes. The default group colours in `plot2` are
either "Okabe-Ito" or "Viridis", depending on the number of groups. But this is
easily changed via the `palette` argument. Note that all palettes supported by
`palette.pals()` and `hcl.pals()` are supported out-of-the-box.^[See the
accompanying help pages of those functions for more details, or read the
[article](https://arxiv.org/pdf/2303.04918.pdf) by Achim Zeileis and Paul
Murrell.] Just pass on an appropriate palette name as a string.

```{r}
plot2(
  Temp ~ Day | Month, airquality,
  type = "l",
  legend.position = "right!",
  palette = "Tableau 10"
)
```

Its possible to add a lot more customization to your plots using the
accompanying `legend.args` and `palette.args` arguments. You can turn off 
the legend title and bounding box, switch the direction of the text, add 
transparency to your colour palette, etc. etc. Here's a quick penultimate
example, where we also show off the enhanced `grid` argument.

```{r}
plot2(
  Temp ~ Day | Month, airquality,
  type = "b", pch = 17,
  grid = grid(), frame.plot = FALSE,
  legend.position = "right!", legend.args = list(bty = "n", title = NULL),
  palette = "Tropic", palette.args = list(alpha = 0.8)
)
```

Finally, because `plot2` is ultimately just a wrapper around base `plot`, any
theming elements set by `par` etc. should carry over too. For nice
out-of-the-box themes, I recommend the **basetheme** package.

```{r}
library(basetheme)
basetheme("royal") # or "clean", "dark", "ink", "brutal", etc.

plot2(
  Temp ~ Day | Month, airquality,
  type = "b", pch = 17,
  grid = grid(), frame.plot = FALSE,
  legend.position = "right!", legend.args = list(bty = "n", title = NULL),
  palette = "Tropic", palette.args = list(alpha = 0.8)
)

basetheme(NULL) # back to default
```