
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%",
  fig.width = 8,
  dpi = 300,
  asp = 0.625
)
```

# plot2

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version/plot2)](https://CRAN.R-project.org/package=plot2)
[![R-universe status badge](https://grantmcdermott.r-universe.dev/badges/plot2)](https://grantmcdermott.r-universe.dev)
[![R-CMD-check](https://github.com/grantmcdermott/plot2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/grantmcdermott/plot2/actions/workflows/R-CMD-check.yaml)
[![Docs](https://img.shields.io/badge/docs-homepage-blue.svg)](https://grantmcdermott.com/plot2/index.html)
<!-- badges: end -->

## What

A lightweight extension of the base R graphics system, with support for
automatic grouping, legends, facets and various other enhancements.

**plot2** is not yet on CRAN, but can be installed from R-universe.

``` r
install.packages("plot2", repos = "https://grantmcdermott.r-universe.dev")
```

Our goal is to submit to CRAN within the first few months of 2024, once we have
settled on some remaining design choices and features support. You can take a 
look at the [open issues](https://github.com/grantmcdermott/plot2/issues) to see
what's currently under consideration. Please feel free to weigh on these if you
have opinions. We want end users to have a say in determining the final product.

## Why

R users are spoiled for choice when it comes to visualization frameworks. The
options include **ggplot2** (arguably the most important graphics
system of the last decade) and **lattice**, not to mention a bewildering array
of extensions built around, on top of, and in between these amazing packages.

As a result, it is perhaps not surprising that the base R graphics system
sometimes gets short shrift. This is unfortunate, because base R offers very
powerful and flexible plotting facilities. Just type `demo(graphics)` or
`demo(persp)` into your R console to get an idea. Or, take a look at
[these](https://github.com/karoliskoncevicius/tutorial_r_introduction/blob/main/baseplotting.md)
[two](https://quizzical-engelbart-d15a44.netlify.app/2021-2022_m2-data-2_visu-2_practice#1)
excellent tutorials. The downside of this power and flexibility is that base R
plotting can require a fair bit of manual tinkering. A case in point is
plotting grouped data with an appropriate legend. Doing so with the generic
`plot()` function can require several function calls or a loop, fiddling with
your plot regions, and then generating the legend manually.

The **plot2** package aims to remove this overhead. It provides a lightweight
(zero dependency) extension of the base R graphics system with various
convenience features, particularly for representing groups with your data. For
example, the core `plot2()` function makes it easy to plot different categories
of a dataset in a single function call and highlight these categories (groups)
using modern colour palettes. Coincident with this grouping support, `plot2()`
also produces automatic legends with scope for further customization. While the
package offers several other enhancements like facets, it tries as far as
possible to be a drop-in replacement for the equivalent base plot function.
Users should generally be able to swap a valid `plot()` call with `plot2()`
without any changes to the expected output.

## Quickstart

The **plot2** website includes a detailed
[introductory tutorial](https://grantmcdermott.com/plot2/vignettes/intro_tutorial.html),
with numerous examples. But here are some quickstart examples of the package in
action.

```{r quickstart1}
library(plot2)
```

Grouped scatterplot with automatic legend:

```{r quickstart2}
# with(iris, plot2(x = Petal.Length, y = Sepal.Length, by = Species)) # atomic
plot2(Sepal.Length ~ Petal.Length | Species, data = iris)             # formula
```

Same plot with a few extra aesthetic tweaks:

```{r quickstart3}
plot2(
  Sepal.Length ~ Petal.Length | Species, 
  data = iris,
  palette = "dark", pch  = 16,
  grid = TRUE, frame = FALSE
)
```

Grouped grouped density plot with automatic legend:

```{r quickstart4}
plot2(
  ~ Petal.Length | Species,
  data = iris,
  type = "density",
  palette = "dark", fill = "by",
  grid = TRUE,
  main = "Distribution of petal lengths by species"
)
```

Grouped scatterplot, combined with facet layout:

```{r quickstart5}
iris2 = transform(iris, Sepals = ifelse(Sepal.Length>6, "Long", "Short"))
plot2(
  Sepal.Length ~ Petal.Length | Sepals, data = iris2,
  facet = ~Species,
  facet.args = list(bg = "grey90"),
  palette = "classic",
  main = "Faceted Sepals!",
  grid = TRUE, frame = FALSE
)
```

Hopefully, these have been enough to pique your interest. Head over to the
[intro tutorial](https://grantmcdermott.com/plot2/vignettes/intro_tutorial.html)
for many more examples, including range plots and customization.