---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "70%",
  fig.width = 8,
  dpi = 300,
  asp = 0.625
)
```

The goal of this intro tutorial is to give you a sense of the main features and
syntax of **plot2**, a lightweight extension of the base R graphics system. We
don't try to cover everything, but you should come away with a good sense of how
the package works and how it can integrate with your own projects.

## Preliminaries

Start by loading the package. For this the examples in this tutorial, we'll be
using a slightly modified version of the `airquality` dataset that comes bundled
with base R. So we'll go ahead and create it now too.

```{r aq}
library(plot2)

aq = airquality
aq$Month = factor(month.abb[aq$Month], levels = month.abb[5:9])
```

## Similarity to `plot()`

As far as possible, `plot2` tries to be a drop-in replacement for regular `plot`
calls.

```{r base_1}
par(mfrow = c(1, 2))

plot(0:10, main = "plot")
plot2(0:10, main = "plot2")

par(mfrow = c(1, 1)) # reset layout
```

Similarly, we can plot elements from a data frame using either the atomic or
formula methods. Here's a simple example using the `aq` dataset that we created
earlier.

```{r plot2_simple}
# with(aq,  plot2(Day, Temp)) # atomic method (same as below)
plot2(Temp ~ Day, data = aq)  # formula method
```

## Grouped data

Where `plot2` starts to diverge from its base counterpart is
with respect to grouped data. In particular, `plot2` allows you to characterize
groups using the `by` argument.^[At this point, experienced base plot users 
might protest that you _can_ colour by groups using the `col` argument, e.g.
`with(aq, plot(Day, Temp, col = Month))`. This is true,
but there are several limitations. First, you don't get an automatic legend.
Second, the base `plot.formula` method doesn't specify the grouping within the
formula itself (not a deal-breaker, but not particularly consistent either).
Third, and perhaps most importantly, this grouping doesn't carry over to line
plots (i.e., type="l"). Instead, you have to transpose your data and use
`matplot`. See
[this](https://stackoverflow.com/questions/10519873/how-to-create-a-line-plot-with-groups-in-base-r-without-loops)
old StackOverflow thread for a longer discussion.]

```{r by}
# plot2(aq$Day, aq$Temp, by = aq$Month) # same as below
with(aq, plot2(Day, Temp, by = Month))
```

An arguably more convenient approach is to use the equivalent formula syntax.
Just place the grouping variable after a vertical bar (i.e., `|`).

```{r formula}
plot2(Temp ~ Day | Month, data = aq)
```

You can use standard base plotting arguments to adjust features of your plot.
For example, change `pch` (plot character) to get filled points.

```{r pch_16}
plot2(
  Temp ~ Day | Month, data = aq,
  pch = 16
)
```

Similarly, converting to a grouped line plot is a simple matter of adjusting the
`type` argument.

```{r type_l}
plot2(
  Temp ~ Day | Month, data = aq,
  type = "l"
)
```

The default behaviour of `plot2` is to represent groups through colour. However,
note that we can automatically adjust `pch` and `lty` by groups too by passing
the `"by"` convenience keyword. This can be used in conjunction with the default
group colouring. Or, as a replacement for group colouring---an option that may
be particularly useful for contexts where colour is expensive or prohibited
(e.g., certain academic journals).

```{r by_lty}
plot2(
  Temp ~ Day | Month, data = aq,
  type = "l",
  col = "black", # override automatic group colours
  lty = "by"     # change line type by group instead
)
```

## Colours

On the subject of group colours, these are easily customized via the `palette`
argument. The default group colours are inherited from the user's default
palette. (Most likely the "R4" set of colors; see `?palette`). However, all of
the various palettes listed by `palette.pals()` and `hcl.pals()` are supported
as convenience strings.^[See the accompanying help pages of those two functions
for more details on the available palettes, or read the
[article](https://arxiv.org/pdf/2303.04918.pdf) by Achim Zeileis and Paul
Murrell.] Note that case-insensitive, partial matching for these convenience
string is allowed. For example:

```{r palette_tableau}
plot2(
  Temp ~ Day | Month, data = aq,
  type = "l",
  palette = "tableau" # or "ggplot2", "okabe-ito", "set2", "harmonic", etc.
)
```

Beyond these convenience strings, users can also supply a valid
palette-generating function for finer control over transparency, colour order,
and so forth. We'll see a demonstration of this further below.

To underscore what we said earlier, colours are inherited from the user's
default palette. So these can also be set globally, just as they can for the
base `plot` function. The next code chunk will set a new default palette for the
remainder of the plots that follow.

```{r global_palette}
# Set the default palette globally via the generic palette function
palette("tableau")
```

## Legend

In all of the preceding plots, you will have noticed that we get an automatic
legend. The legend position and look can be customized with the `legend`
argument. At a minimum, you can pass the familiar legend position keywords as a
convenience string ("topright", "bottom", "left", etc.). Moreover, a key feature
of `plot2` is that we can easily and elegantly place the legend _outside_ the
plot area by adding a trailing "!" to these keywords. (As you may have realised,
the default legend position is "right!".) Let's demonstrate by moving the legend
to the left of the plot:

```{r legend_bottom}
plot2(
  Temp ~ Day | Month, data = aq,
  type = "l",
  legend = "left!"
)
```

Beyond the convenience of these positional keywords, the `legend` argument also
permits additional customization by passing an appropriate function (or, a list
of arguments that will be passed on to the standard `legend()` function
internally.) So you can change or turn off the legend title, remove the bounding
box, switch the direction of the legend text to horizontal, etc. Here's a
grouped density plot example, where we also add some shading by specifying that
the background colour should vary by groups too.

```{r density_topright}
with(
  aq,
  plot2(
    density(Temp),
    by = Month,
    fill = "by",                         # add fill by groups
    grid = TRUE,                         # add background grid
    legend = list("topright", bty = "o") # change legend features
  )
)
```

## Interval plots

`plot2` adds supports for interval plots via the `"pointrange"`, `"errorbar"`,
`"ribbon"` type arguments. An obvious use-case is for regression analysis and
prediction.

```{r ribbon_pred}
mod = lm(Temp ~ 0 + Month / Day, data = aq)
aq = cbind(aq, predict(mod, interval = "confidence"))

with(
  aq,
  plot2(
    x = Day, y = fit,
    ymin = lwr, ymax = upr,
    by = Month,
    type = "ribbon",
    grid = TRUE,
    main = "Model predictions"
  )
)
```

Similarly, we can grab the model estimates to produce nice coefficient plots.

```{r pointrange, warning = FALSE}
coeftab = data.frame(
  gsub("Month", "", names(coef(mod))),
  coef(mod),
  confint(mod)
) |>
  setNames(c("term", "estimate", "ci_low", "ci_high"))

with(
  subset(coeftab, !grepl("Day", term)),
  plot2(
    x = term, y = estimate,
    ymin = ci_low, ymax = ci_high,
    type = "pointrange", # or: "errobar", "ribbon"
    pch = 19, col = "dodgerblue",
    grid = TRUE,
    main = "Average Monthly Effect on Temperature"
  )
)
```

## Facets

Alongside the standard "by" grouping approach that we have seen thus far,
**plot2** also supports faceted plots. Mirroring the main `plot2` function, the
`facet` argument accepts both atomic and formula methods.

```{r facet_simple}
with(
  aq,
  plot2(
    x = Day, y = fit,
    ymin = lwr, ymax = upr,
    type = "ribbon",
    facet = Month, ## <- facet, not by
    grid = TRUE,
    main = "Predicted air temperatures"
  )
)
```

By default, facets will be arranged in a square configuration if more than
three facets are detected. Users can override this behaviour by supplying
`nrow` or `ncol` in the "facet.args" helper function. (The margin padding
between individual facets can also be adjusted via the `fmar` argument.) Note
that we can also reduce axis label redundancy by turning off the plot frame.

```{r facet_nrow}
with(
  aq,
  plot2(
    x = Day, y = fit,
    ymin = lwr, ymax = upr,
    type = "ribbon",
    facet = Month,
    facet.args = list(nrow = 1),
    grid = TRUE, frame = FALSE,
    main = "Predicted air temperatures"
  )
)
```

Here's a slightly fancier version where we combine facets with (by) colour
grouping, add a background fill to the facet text, and also add back the
original values to our model predictions. For this particular example, we'll use
the `facet = "by"` convenience shorthand to facet along the same month variable
as the colour grouping. But you can easily specify different `by` and `facet`
variables if that's what your data support.

```{r facet_fancy}
# Plot the original points 
with(
  aq,
  plot2(
    x = Day, y = Temp,
    by = Month,
    facet = "by", facet.args = list(bg = "grey90"),
    palette = "dark2",
    grid = TRUE, frame = FALSE, ylim = c(50, 100),
    main = "Actual and predicted air temperatures"
  )
)
# Add the model predictions to the same plot 
with(
  aq,
  plot2(
    x = Day, y = fit,
    ymin = lwr, ymax = upr,
    by = Month, facet = "by",
    type = "ribbon",
    palette = "dark2",
    add = TRUE
  )
)
```

Again, the `facet` argument also accepts a formula interface. One particular use
case is for two-sided formulas, which arranges the facet layout in a fixed grid
arrangement. Here's a simple (if contrived) example.

```{r facet_grid}
aq$hot = ifelse(aq$Temp>=75, "hot", "cold")
aq$windy = ifelse(aq$Wind>=15, "windy", "calm")

plot2(
 Temp ~ Day, data = aq,
 facet = windy ~ hot,
 # the rest of these arguments are optional...
 facet.args = list(col = "white", bg = "black"),
 pch = 16, col = "dodgerblue",
 grid = TRUE, frame = FALSE, ylim = c(50, 100),
 main = "Daily temperatures vs. wind"
)
```

## Customization

Customizing your plots further is straightforward, whether that is done by
changing global parameters (via `par`) or invoking `plot2` arguments. Here's a
quick penultimate example, where we change our point character, tick labels, and
font family globally, before adding some transparency to our colour palette, and
use Tufte-style floating axes with a background panel grid.

```{r hershey_plus}
par(
  pch    = 16,           # Filled points as default
  las    = 1,            # Horizontal axis tick labels
  family = "HersheySans" # Use a (built-in) Hershey font instead of Arial default
)

plot2(
  Temp ~ Day | Month, data = aq,
  type = "b",
  palette = palette.colors(palette = "tableau", alpha = 0.5),
  frame = FALSE, grid = TRUE,
  main = "Daily temperatures by month"
)
```

(For access to a wider array of fonts, you might consider the
**showtext** package
([link](https://cran.r-project.org/web/packages/showtext/vignettes/introduction.html)).)

At the risk of repeating ourselves, the use of `par` in the previous example
again underscores the correspondence with the base graphics system. Because
`plot2` is effectively a convenience wrapper around base `plot`, any global
elements that you have set for the latter should carry over to the former. For
nice out-of-the-box themes, we recommend the **basetheme** package
([link](https://github.com/karoliskoncevicius/basetheme)).

```{r basethme_royal}
par(pch = 15, las = 0, family = "") # change/revert global changes from above

library(basetheme)
basetheme("royal") # or "clean", "dark", "ink", "brutal", etc.

plot2(
  Temp ~ Day | Month, data = aq,
  type = "b",
  pch = "by",
  palette = "tropic",
  main = "Daily temperatures by month"
)

basetheme(NULL)  # back to default theme
dev.off()
```

## Conclusion

In summary, consider the **plot2** package if you are looking for base R `plot`
functionality with some added convenience features. You can use pretty much the
same syntax and all of your theme elements should carry over too. It has no
dependencies other than base R itself and this makes it an attractive option for
package developers or situations where dependency management is expensive (e.g.,
an R application running in a browser via
[WebAssembly](https://docs.r-wasm.org/webr/latest/)).

