---
title: "Introduction to tinyplot"
format: html
engine: knitr
vignette: |
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "70%",
  fig.width = 8,
  dpi = 300,
  asp = 0.625
)
```

The goal of this intro tutorial is to give you a sense of the main features and
syntax of **tinyplot**, a lightweight extension of the base R graphics system.
We don't try to cover everything, but you should come away with a good
understanding of how the package works and how it can integrate with your own
projects.

We start this tutorial by loading the package and a slightly modified version
of the `airquality` dataset that comes bundled with base R.

```{r aq}
library(tinyplot)

aq = transform(
  airquality,
  Month = factor(month.abb[Month], levels = month.abb[5:9]),
  hot   = ifelse(Temp>=75, "hot", "cold"),
  windy = ifelse(Wind>=15, "windy", "calm")
)
```

## Equivalence with `plot()`

As far as possible, `tinyplot` tries to be a drop-in replacement for regular `plot`
calls.

```{r base_1}
par(mfrow = c(1, 2))

plot(0:10, main = "plot")
tinyplot(0:10, main = "tinyplot")

par(mfrow = c(1, 1)) # reset layout
```

Similarly, we can plot elements from a data frame using either the atomic or
formula methods. Here's a simple example using the `aq` dataset that we created
earlier.

```{r tinyplot_simple}
# with(aq,  tinyplot(Day, Temp)) # atomic method (same as below)
tinyplot(Temp ~ Day, data = aq)  # formula method
```

::: {.callout-tip}
## Tip: `plt()` shorthand alias

Use the `plt()` alias instead of `tinyplot()` to save yourself a few keystrokes.
For example:

```{r plt_simple}
#| eval: false
plt(Temp ~ Day, data = aq)
```

Feel free to try this shorthand with any of the plots that follow; `tinyplot(<args>)`
and `plt(<args>)` should produce identical results.
:::

## Grouped data

Where `tinyplot` starts to diverge from its base counterpart is
with respect to grouped data. In particular, `tinyplot` allows you to characterize
groups using the `by` argument.^[At this point, experienced base plot users 
might protest that you _can_ colour by groups using the `col` argument, e.g.
`with(aq, plot(Day, Temp, col = Month))`. This is true,
but there are several limitations. First, you don't get an automatic legend.
Second, the base `plot.formula` method doesn't specify the grouping within the
formula itself (not a deal-breaker, but not particularly consistent either).
Third, and perhaps most importantly, this grouping doesn't carry over to line
plots (i.e., type="l"). Instead, you have to transpose your data and use
`matplot`. See
[this](https://stackoverflow.com/questions/10519873/how-to-create-a-line-plot-with-groups-in-base-r-without-loops)
old StackOverflow thread for a longer discussion.]

```{r by}
# tinyplot(aq$Day, aq$Temp, by = aq$Month) # same as below
with(aq, tinyplot(Day, Temp, by = Month))
```

An arguably more convenient approach is to use the equivalent formula syntax.
Just place the "by" grouping variable after a vertical bar (i.e., `|`).

```{r formula}
tinyplot(Temp ~ Day | Month, data = aq)
```

You can use standard base plotting arguments to adjust features of your plot.
For example, change `pch` (plot character) to get filled points and `cex`
(character expansion) to change their size.

```{r pch_16}
tinyplot(
  Temp ~ Day | Month, data = aq,
  pch = 16,
  cex = 2
)
```

Similarly, converting to a grouped line plot is a simple matter of adjusting the
`type` argument.

```{r type_l}
tinyplot(
  Temp ~ Day | Month, data = aq,
  type = "l"
)
```

The default behaviour of `tinyplot` is to represent groups through colour. However,
note that we can automatically adjust `pch` and `lty` by groups too by passing
the `"by"` convenience keyword. This can be used in conjunction with the default
group colouring. Or, as a replacement for group colouring---an option that may
be particularly useful for contexts where colour is expensive or prohibited
(e.g., certain academic journals).

```{r by_lty}
tinyplot(
  Temp ~ Day | Month, data = aq,
  type = "l",
  col = "black", # override automatic group colours
  lty = "by"     # change line type by group instead
)
```

The `"by"` convenience argument is also available for mapping group colours to
background fill `bg` (alias `fill`). One use case is to override the grouped
border colours for filled plot characters and instead pass them through the
background fill.

```{r}
tinyplot(
  Temp ~ Day | Month, data = aq,
  pch = 21,      # use filled circles
  col = "black", # override automatic group (border) colours of points
  fill = "by"    # use background fill by group instead
)
```

## Colours

On the subject of group colours, the default palette should adjust automatically
depending on the class and cardinality of the grouping variable. For example, a
sequential ("viridis") palette will be used if an ordered factor is detected.

```{r palette_ordered}
tinyplot(
  Temp ~ Day | ordered(Month), data = aq,
  pch = 16
)
```

However, this behaviour is easily customized via the `palette` argument.
The default set of discrete colours are inherited from the user's current global
palette. (Most likely the "R4" set of colors; see `?palette`). However, all of
the various palettes listed by `palette.pals()` and `hcl.pals()` are supported
as convenience strings.^[See the accompanying help pages of those two functions
for more details on the available palettes, or read Zeileis & Murrell
(2023, The R Journal, [doi:10.32614/RJ-2023-071](https://doi.org/10.32614/RJ-2023-071)).]
Note that case-insensitive, partial matching for these convenience
string is allowed. For example:

```{r palette_tableau}
tinyplot(
  Temp ~ Day | Month, data = aq,
  type = "l",
  palette = "tableau" # or "ggplot", "okabe", "set2", "harmonic", etc.
)
```

Beyond these convenience strings, users can also supply a valid
palette-generating function for finer control and additional options.^[For
example, if you have installed the **ggsci** package
([link](https://nanx.me/ggsci/index.html)) then you could use
`palette = ggsci::pal_npg()` to generate a palette consistent with those used by
the Nature Publishing Group.] You can also use the `alpha` argument to adjust
the (alpha) transparency of your colours:

```{r}
tinyplot(
  Temp ~ Day | Month, data = aq,
  pch = 19, cex = 2,
  palette = "tableau",
  alpha = 0.3
)
```

To underscore what we said earlier, colours are inherited from the user's
current palette. So these can also be set globally, just as they can for the
base `plot` function. The next code chunk will set a new default palette for the
remainder of the plots that follow.

```{r global_palette}
# Set the default palette globally via the generic palette function
palette("tableau")
```

## Legend

In all of the preceding plots, you will have noticed that we get an automatic
legend. The legend position and look can be customized with the `legend`
argument. At a minimum, you can pass the familiar legend position keywords as a
convenience string ("topright", "bottom", "left", etc.). Moreover, a key feature
of `tinyplot` is that we can easily and elegantly place the legend _outside_ the
plot area by adding a trailing "!" to these keywords. (As you may have realised,
the default legend position is "right!".) Let's demonstrate by moving the legend
to the left of the plot:

```{r legend_bottom}
tinyplot(
  Temp ~ Day | Month, data = aq,
  type = "l",
  legend = "left!"
)
```

Beyond the convenience of these positional keywords, the `legend` argument also
permits additional customization by passing an appropriate function (or, a list
of arguments that will be passed on to the standard `legend()` function
internally.) So you can change or turn off the legend title, remove the bounding
box, switch the direction of the legend text to horizontal, etc. Here's a
grouped density plot example, where we also add some shading by specifying that
the background colour should vary by groups too.

```{r density_topright}
tinyplot(
  ~ Temp | Month,
  data = aq,
  type = "density",
  fill = "by",                         # add fill by groups
  grid = TRUE,                         # add background grid
  legend = list("topright", bty = "o") # change legend features
)
```

All of the legend examples that we have seen thus far are representations of
discrete groups. However, please note that **tinyplot** also supports grouping
by continuous variables, which automatically yield gradient legends.

```{r legend_gradient}
tinyplot(Temp ~ Wind | Ozone, data = aq, pch = 19)
```

Gradient legends (and plots) can be customized in an identical manner to
discrete legends by adjusting the keyword positioning, palette choice, alpha
transparency etc. Here is a quick adaptation of the previous plot to
demonstrate. Note that here we pass a special convenience argument to
`bg`/`fill`; if it detects a numeric in the range of `[0,1]`, then it
automatically inherits the grouped colour mappings but with added transparency.

```{r legend_gradient2}
tinyplot(
  Temp ~ Wind | Ozone, data = aq,
  pch  = 21,      # use filled plot character
  cex  = 2,
  col  = "black", # override automatic (grouped) border colour of points
  fill = 0.5,     # use background fill instead with added alpha transparency 
)
```

## More plot types

We have already seen several plot types above such as `"p"` (points), `"l"`
(lines), and `"density"`. In general, **tinyplot** supports all of the primitive
plot types/elements available in base R, as well as a number of additional plot
types that can be a bit tedious to code up manually. The full list of supported
plot types can be viewed in
[this pinned GitHub issue](https://github.com/grantmcdermott/tinyplot/issues/97),
or by checking the
[`?tinyplot`](https://grantmcdermott.com/tinyplot/man/tinyplot.html#arguments)
documentation.

For example, **tinyplot** supports interval plots via the `"pointrange"`,
`"errorbar"`, `"ribbon"`, and related type arguments. A canonical use-case is
coefficient plots: 

```{r pointrange, warning = FALSE}
mod = lm(Temp ~ 0 + Month / Day, data = aq)

# grab coefs of interest
monthcoefs = data.frame(
  gsub("Month", "", names(coef(mod))),
  coef(mod),
  confint(mod)
  ) |>
  setNames(c("term", "estimate", "ci_low", "ci_high")) |>
  subset(!grepl("Day", term))

# plot
with(
  monthcoefs,
  tinyplot(
    x = term, y = estimate,
    ymin = ci_low, ymax = ci_high,
    type = "pointrange", # or: "errobar", "ribbon"
    pch = 19, col = "dodgerblue",
    grid = TRUE,
    main = "Average Monthly Effect on Temperature"
  )
)
```

**tinyplot** also supports special types to fit models and display their
predictions, along with confidence intervals. Here is a somewhat silly example
where we fit a linear model to predict temperature by day of month.^[The grouped
setting here makes this visualization equivalent to
`predict(lm(Temp ~ 0 + Month / Day, data = aq), interval = "confidence")`.]

```{r ribbon_pred}
tinyplot(
  Temp ~ Day | Month, aq,
  type = "lm",
  grid = TRUE,
  main = "Linear model"
)
```

The default behaviour of these model types can be adjusted by passing
appropriate arguments to the equivalent _functional_ version of the type in
question. These functional types all follow a `type_<typename>` syntax, so that
`"lm"` is paired with `type_lm()`, etc. Below we illustrate an adapted
generalised linear model, where we passing the binomial
[family](https://search.r-project.org/R/refmans/stats/html/family.html) argument
and thus fit a logistic regression.

```{r ribbon_pred_glm}
tinyplot(
  I(Temp > 80) ~ Wind, aq,
  type = type_glm(family = binomial),
  main = "Logit model: Temps above 80 °F"
)
```

We will see examples of more plot types below, including other model prediction
types. Please also take a look at
[the dedicated Plot Types vignette](types.qmd) for explicit details about all of
the different plot types that **tinyplot** supports, as well as how to create
your own custom types.

## Facets

Alongside the standard "by" grouping approach that we have seen thus far,
**tinyplot** also supports faceted plots. Mirroring the main `tinyplot`
function, the `facet` argument accepts both atomic and formula methods. In
general, however, we recommend the formula version as being safer since it does
a better job of handling missing values.

```{r facet_simple}
tinyplot(
  Temp ~ Day, aq,
  facet = ~Month, ## <= facet, not by
  type = "lm",
  grid = TRUE,
  main = "Predicted air temperatures"
)
```

Facets are easily combined with grouping. This can either be done separately
(i.e., distinct arguments for `by` and `facet`), or along the same dimension.
For the latter case, we provide a special `facet = "by"` convenience shorthand.

```{r facet_by}
tinyplot(
  Temp ~ Day | Month, aq,
  facet = "by", # facet along same dimension as groups
  type = "lm",
  grid = TRUE,
  main = "Predicted air temperatures"
)
```

To customize facets, simply pass a list of named arguments through the companion
`facet.args` argument. Customization options include: override the default
"square" facet window arrangement; allow free-scaled axes so that the limits of
each individual facet are drawn independently; adjust the padding (margin)
between individual facets; change the facet title text and background; etc.
Here is a simple example where we (1) arrange the facets in a single row, (2)
add some background fill to the facet text, and (3) and reduce axis redundancy
by turning off the plot frame.

```{r facet_nrow}
tinyplot(
  Temp ~ Day, aq,
  facet = ~Month, facet.args = list(nrow = 1, bg = "grey90"),
  type = "lm",
  grid = TRUE,
  frame = FALSE, # turning off the plot frame means only outer axes will be printed
  main = "Predicted air temperatures"
)
```

The `facet.args` customizations can also be set globally via the `tpar()`
function. We will revisit this idea in the [Themes](#themes) section below.

Finally, the `facet` argument also accepts a _two-sided_ formula for arranging
facets in a fixed grid layout. Here's a simple (if contrived) example.

```{r facet_grid}
tinyplot(
 Temp ~ Day, data = aq,
 facet = windy ~ hot,
 # the rest of these arguments are optional...
 facet.args = list(col = "white", bg = "black"),
 pch = 16, col = "dodgerblue",
 grid = TRUE, frame = FALSE, ylim = c(50, 100),
 main = "Daily temperatures vs. wind"
)
```

## Layers

In many contexts, it is convenient to build plots step-by-step, adding layers
with different elements on top of a base. The **tinyplot** package offers a few
ways to achieve this layering effect.

Similar to many base plotting functions, users can invoke the
`tinyplot(..., add=TRUE)` argument to draw a plot on top of an existing one,
rather than opening a new window. However, while this argument is useful, it can
become verbose since it requires users to make very similar successive calls,
with many shared arguments.

For this reason, the **tinyplot** package provides a special
`tinyplot_add()` convenience function for adding layers to
an existing tinyplot. The idea is that users need simply pass the _specific_
arguments that they want to add or modify relative to the base layer, and all
arguments will be inherited from the original.

An example may help to demonstrate. Here we first draw some faceted points with
group colouring and various other aesthetic tweaks. Next, we add regression
fits with a simple `tinyplot_add(type = "lm")` call. Notice that the original
grouping, data and aesthetic options are all carried over correctly, without
having to repeat ourselves.

```{r tinyplot_add}
tinyplot(
  Temp ~ Day | Month, aq,
  facet = "by", facet.args = list(bg = "grey90"),
  palette = "dark2",
  legend = FALSE,
  grid = TRUE,
  axes = "l",
  ylim = c(50, 100),
  main = "Actual and predicted air temperatures"
)
# Add regression fits
tinyplot_add(type = "lm")
```

::: {.callout-tip}
## Tip: `plt_add()` shorthand alias

Much like `plt()` is an alias for `tinyplot()`, you can save yourself a few
keystrokes by typing `plt_add()` instead of `tinyplot_add()`:

```{r plt_add}
#| eval: false
plt_add(type = "lm")
```
:::

A related---but distinct---concept to adding plot layers is _drawing_ on a plot.
The canonical use case is annotating your plot with text or some other
function-based (rather than data-based) logic. For example, you may want to
demarcate some threshold values with horizontal or vertical lines, or simply
annotate your plot with text. The **tinyplot** way to do this is by passing the
`tinyplot(..., draw = <draw_function>)` argument. Here we demonstrate with a
simplified version of our facet grid example from earlier.

```{r draw_simple}
tinyplot(
 Temp ~ Day, data = aq,
 facet = windy ~ hot,
 # draw a horizontal (threshold) line in each facet using the abline function
 draw = abline(h = 75, lty = 2, col = "hotpink")
)
```

Compared to "manually" drawing these elements on a plot _ex post_---e.g., via a
separate `abline()` call---there are several advantages to the idiomatic
**tinyplot** interface. First, the `draw` argument is facet-aware and will
ensure that each individual facet is correctly drawn upon. Second, you can
leverage the special `ii` internal counter to draw programmatically across
facets (see
[here](https://github.com/grantmcdermott/tinyplot/pull/245#issue-2642589951)).
Third, the `draw` argument is fully generic and accepts _any_
drawing/annotating function. You can combine multiple drawing functions by
wrapping them with curly brackets`{}`, and even pass `tinyplot()` back towards
itself.

Here's a slightly more complicated "spaghetti" plot example, where we pass
multiple functions through `draw = {...}`, including drawing all of the
lines in the background (of each facet) via a secondary `tinyplot()` call.

```{r draw_spaghetti}
tinyplot(
  Temp ~ Day | Month, aq, facet = "by", lwd = 3, type = "l",
  frame = FALSE, legend = FALSE, ylim = c(50, 100),
  draw = {
    tinyplot(Temp ~ Day | Month, aq, col = "grey", type = "l", add = TRUE)
    abline(h = 75, lty = 2)
    text(5.5, 75, "Cold", pos = 1, offset = 0.4)
    text(5.5, 75, "Hot", pos = 3, offset = 0.4)
  }
)
```

## Themes

Customizing your plots further is straightforward, whether that is done directly
by changing `tinyplot` arguments for a single plot, or by setting global
parameters. For setting global parameters, users can invoke the standard `par()`
arguments. But for improved convenience and integration with the rest of the
package, we recommend that users instead go via
[`tpar()`](https://grantmcdermott.com/tinyplot/man/tpar.html), which is an
extended version of `par()` that supports all of the latter's parameters plus
some `tinyplot`-specific ones. Here's a quick penultimate example, where we
impose several global changes (e.g., rotated axis labels, removed plot frame to
get Tufte-style floating axes, etc.) before drawing the plot. 
change our point character, tick labels, and font family globally, before adding
some transparency to our colour palette, and use Tufte-style floating axes with
a background panel grid.

```{r hershey_plus}
op = tpar(
  bty    = "n",           # No box (frame) around the plot 
  family = "HersheySans", # Use R's Hershey font instead of Arial default
  grid   = TRUE,          # Add a background grid
  las    = 1,             # Horizontal axis tick labels
  pch    = 16             # Filled points as default
)

tinyplot(
  Temp ~ Day | Month, data = aq,
  type = "b",
  alpha = 0.5,
  main = "Daily temperatures by month"
)
```

_Note: For access to a much wider variety of fonts, you might consider the
**showtext** package
([link](https://cran.r-project.org/web/packages/showtext/vignettes/introduction.html))._

At the risk of repeating ourselves, the use of `(t)par` in the previous example
again underscores the correspondence with the base graphics system. Because
`tinyplot` is effectively a convenience wrapper around base `plot`, any global
elements that you have set for the latter should carry over to the former. For
nice out-of-the-box themes, we recommend the **basetheme** package
([link](https://github.com/karoliskoncevicius/basetheme)).

```{r basethme_royal}
tpar(op) # revert global changes from above

library(basetheme)
basetheme("royal") # or "clean", "dark", "ink", "brutal", etc.

tpar(pch = 15) # filled squares as first pch type

tinyplot(
  Temp ~ Day | Month, data = aq,
  type = "b",
  pch = "by",
  palette = "tropic",
  main = "Daily temperatures by month"
)

basetheme(NULL)  # back to default theme
dev.off()
```

## Saving plots

A final point to note is that **tinyplot** offers convenience features for
exporting plots to disk. Simply invoke the `file` argument to specify the
relevant file path (including the extension type). You can customize the output
dimensions (in inches) via the accompanying `width` and `height`
arguments.^[The default dimensions are 7x7, with a resolution of 300 DPI.
However, these too can be customized via the `file.width`, `file.height`, and
`file.res` parameters in
[`tpar()`](https://grantmcdermott.com/tinyplot/man/tpar.html).]

```{r save_plot}
#| eval: false
tinyplot(
  Temp ~ Day | Month, data = aq,
  file = "aq.png", width = 8, height = 5
)

# optional: delete the saved plot
unlink("aq.png")
```

Alongside convenience, the benefit of this native **tinyplot** approach (versus
the traditional approach of manually opening an external graphics device, e.g.
`png()`) is that all of your current graphic settings are automatically carried
over to the exported file. Feel free to try yourself by setting some global
graphics parameters via `tpar()` and then using `file` to save a plot.

## Conclusion

In summary, consider the **tinyplot** package if you are looking for base R `plot`
functionality with added convenience features. You can use (nearly) the exact
same syntax and all of your theme elements should carry over too. It has no
dependencies other than base R itself and this should make it an attractive
option for package developers, as well as situations where dependency management
is expensive (e.g., production pipelines, continuous integration or an R
application running in a browser via
[WebAssembly](https://docs.r-wasm.org/webr/latest/)).

