---
title: "Themes"
---

# `par()` and `tpar()`

In base R, the appearance of plots can be controlled using graphical parameters defined with the `par()` function. **tinyplot** offers an extended version of this function---`tpar()`---with additional visual parameters. For example, we can change the alignment of axis labels using the `adj.xlab` and `adj.ylab` parameters.

```{r}
library(tinyplot)
tinyplot(mpg ~ hp, data = mtcars)

tpar("adj.ylab" = 1, "adj.xlab" = 1)
tinyplot(mpg ~ hp, data = mtcars)
```

See `?par()` and `?tpar()` for a list of available parameters.

# `tinytheme()`

Instead of modifying every graphical parameter individually, users can call `tinytheme()` to set multiple parameters at once, according to pre-defined styles. One of the built-in styles is called "ipsum", inspired by the [**hrbrthemes** for **ggplot2**.](https://cran.r-project.org/web/packages/hrbrthemes/index.html)

```{r}
library(tinyplot)

tinytheme("ipsum")

tinyplot(Sepal.Width ~ Sepal.Length | Species,
  facet = ~Species,
  data = iris,
  pch = 16,
  main = "Title of this plot",
  sub = "A smaller subtitle",
  palette = "okabeito")
```

The changes are persistent, and will apply to all graphs until the graphical device is closed.

```{r}
tinyplot(Sepal.Width ~ Sepal.Length | Species,
  facet = ~Species,
  data = iris,
  pch = 16,
  main = "Title of this plot",
  sub = "A smaller subtitle",
  palette = "okabeito")
```

```{r}
#| eval: false
tinyplot(mpg ~ hp, data = mtcars, main = "Cars")
```





