---
title: "Themes"
format: html
engine: knitr
vignette: |
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "70%",
  fig.width = 8,
  dpi = 300,
  asp = 0.625
)
```

The base R graphics aesthetic tends to divide option. Some people like the
default minimalist look of base R plots and/or are happy to customize the (many)
graphical parameters that are available to them. Others find base R plots ugly
and don't want to spend time endlessly tweaking different parameters. Similarly,
the inherent "canvas" approach to drawing base R graphics, with fixed placement
for plot elements, means that plots don't adjust dynamically and this can lead
to awkward whitespace artifacts unless the user explicitly accounts for them.

Regardless of where you stand in this debate, the **tinyplot** view is that base
R graphics should ideally combine flexibility and ease of use with aesthetically
pleasing end results. One way that we enable this is via _themes_.

## `tinytheme()`

The `tinytheme()` function provides a mechansim for easily changing the look of
your plots to match a variety of pre-defined styles. Behind the scenes, this
works by simultaneously setting a group of graphical parameters to achieve a
particular aesthetic. Let's take a look at the "minimal" theme for example,
which is inspired by a well-known **ggplot2** theme.

```{r}
library(tinyplot)

tinytheme("minimal")

tinyplot(
  Sepal.Width ~ Sepal.Length | Species,
  facet = "by",
  data = iris,
  main = "Title of the plot",
  sub = "A smaller subtitle"
)
```

One particular feature that may interest users is the fact that `tinytheme()`
uses some internal ~~magic~~ logic to dynamically adjust plot margins to avoid
whitespace. For example, when long horizontal y-axis labels are detected:

```{r}
tinyplot(
  I(Sepal.Width*1e5) ~ Sepal.Length | Species,
  facet = "by",
  data = iris,
  main = "Title of the plot",
  sub = "The left-margin adjusts to accomodate the long y-axis labels"
)
```

As you may have noticed, the changes made by `tinytheme()` are persistent, and
apply to all subsequent `tinyplot()` calls.

```{r}
tinyplot(mpg ~ hp, data = mtcars, main = "Fuel efficiency vs. horsepower")
```

To reset graphical parameters to factory defaults, call  `tinytheme()` without arguments.

```{r}
tinytheme()
tinyplot(mpg ~ hp, data = mtcars, main = "Fuel efficiency vs. horsepower")
```


### Gallery

The `tinytheme()` function ships with a number of built-in themes, which build
on top of each other in an approximately hierarchical manner (adding or
subtracting elements from the predecessor theme).

```{r}
p = function() {
  tinyplot(
    Sepal.Width ~ Sepal.Length | Species,
    facet = "by",
    data = iris,
    main = "Title of the plot",
    sub = "A smaller subtitle"
  )
}
```


<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| fig.keep: "all" -->
<!-- thms = eval(formals(tinytheme)$theme) -->

<!-- for (thm in thms) { -->
<!--   tinytheme(thm) -->
<!--   tinyplot( -->
<!--     Sepal.Length ~ Petal.Length | Species, facet = "by", data = iris, -->
<!--     main = paste0('tinytheme("', thm, '")'), -->
<!--     sub = "Subtitle of the plot" -->
<!--   ) -->
<!-- } -->

<!-- # Reset -->
<!-- tinytheme() -->
<!-- ``` -->

#### default

```{r}
tinytheme() ## same as tinytheme("default")
p()
```

#### basic

```{r}
tinytheme("basic")
p()
```

#### clean

```{r}
tinytheme("clean")
p()
```

#### clean2

```{r}
tinytheme("clean2")
p()
```

#### bw

```{r}
tinytheme("bw")
p()
```

#### classic

```{r}
tinytheme("classic")
p()
```

#### minimal

```{r}
tinytheme("minimal")
p()
```

#### ipsum

```{r}
tinytheme("ipsum")
p()
```

#### dark

```{r}
tinytheme("dark")
p()
```

#### tufte

```{r}
tinytheme("tufte")
p()
```

#### void

```{r}
tinytheme("void")
p()
```


```{r}
# Reset to default theme
tinytheme()
```

Please feel free to make suggestions about themes, or contribute new themes by
[opening a Pull Request on Github.](https://github.com/grantmcdermott/tinyplot)

### Custom themes

Tweaking existing themes is easy. For example, the `tinytheme()` function also
accepts any graphical parameter supported by `tpar()`/`par()` and applies them
in persistent fashion.

```{r}
#| layout-ncol: 2
tinytheme("ipsum", pch = 2, col.axis = "red", cex.main = 2, family = "Palatino")
tinyplot(mpg ~ hp, data = mtcars, main = "Fuel efficiency vs. horsepower")
tinyplot(hp ~ mpg, data = mtcars, main = "Horsepower vs. fuel efficiency")
```

```{r}
# reset
tinytheme()
```

::: {.callout-tip}
## Font families

Fonts are a suprisingly effective way to add impact and personality to your
plots. While it is perhaps underappreciated, base R actually comes with built-in
support for quite a few font families. In the code chunk above we used the
well-known "Palatino" font, but support also extends to other popular LaTeX and
PDF fonts (see `?pdfFonts`), as well as custom fonts like the Hershey family
(see `?Hershey`).

For access to a _much_ wider variety of fonts, you might consider the excellent
**showtext** package
([link](https://cran.r-project.org/package=showtext)).
This package allows you to install _any_ font family from the Google Font
catalog, either on-the-fly or downloaded to your permanent fontbook collection.
It plays very nicely with **tinyplot**.

:::

Similarly, to create your own themes "from scratch", set the theme to
`"default"` and passing additional graphical parameters to `tinytheme()`.

```{r}
tinytheme("default", font.main = 3, col = "red")
p()
```

```{r}
# Reset to default theme
tinytheme()
```

::: {.callout-tip}
To see the full list of parameters that defines a particular theme, simply
assign them to an object. This can be helpful if you want to explore creating
your own custom theme, or tweak an existing theme.

```{r}
#| eval: false

# parms = tinytheme("clean")   # assigns the theme at the same time
parms = tinyplot:::theme_clean # doesn't assign the theme

# show the list of parameters used in the "clean" theme
parms
```

:::

## Manual customization with `tpar()`

_Subtitle: And comparison with `par()`_

Themes are a powerful and convenient way to customize your plots. But they are
not the only game in town. As any base R plotter would tell you, another way
to customize your plots by setting global graphics parameters via
[`par()`](https://search.r-project.org/R/refmans/graphics/html/par.html).
If you prefer this approach, then the good news is that it is fully compatable
with **tinyplot**.^[After all, a tinyplot is just a base plot with added
convenience features. We still use the same **graphics** engine under the hood
and any settings and workflows for `plot()` should (ideally) carry over to
`tinyplot()` too.] However, we recommend that you rather use
[`tpar()`](https://grantmcdermott.com/tinyplot/man/tpar.html),
which is an extended version of `par()` that supports all of the latter's
parameters plus some `tinyplot`-specific upgrades.

Here is a quick example, where we impose several global changes (e.g., change
the font family, use Tufte-style floating axes with a background panel grid,
rotate tick labels, etc.) before drawing the plot.

```{r hershey_plus}
op = tpar(
  bty    = "n",           # No box (frame) around the plot 
  family = "HersheySans", # Use R's Hershey font instead of Arial default
  grid   = TRUE,          # Add a background grid
  las    = 1,             # Horizontal axis tick labels
  pch    = 19             # Larger filled points as default
)

tinyplot(Sepal.Length ~ Petal.Length | Species, data = iris, alpha = 0.5)

# optional: reset to the original parameter settings
tpar(op)
```

Again, this approach should feel very familiar to experience base R plotters.
But we will drive home the point by exploring one final difference between
vanilla `par()` and the enhanced `tpar()` equivalent...

The graphical parameters set by `par()` stay in force as long as a graphical
device stays open. On the other hand, these parameters are reset when the
plotting window is closed or, for example, when executing a new code chunk in a
Quarto notebook.^[The **knitr** package, which provides the rendering engine for
Quarto and R Markdown, also has a `global.par` option to overcome this
limitation. See [here](https://yihui.org/knitr/options/#package-options).]

```{r}
par(col = "red", pch = 4)
tinyplot(mpg ~ hp, data = mtcars)
```
```{r}
tinyplot(wt ~ qsec, data = mtcars)
```

In contrast, graphical parameters set by `tpar()` can persist
across devices and Quarto code chunks thanks to a built-in "hook" mechanism (see
`?setHook`). To enable this persistence, we must invoke the `hook=TRUE`
argument.

```{r}
tpar(col = "red", pch = 4, hook = TRUE)
tinyplot(mpg ~ hp, data = mtcars)
```
```{r}
tinyplot(wt ~ qsec, data = mtcars)
```

```{r}
# reset defaults
tinytheme()
```

(Fun fact: Behind the scenes, `tinytheme(<theme_name>)` is simply passing a list
of parameters to `tpar(..., hook = TRUE)`.)
