---
title: "Tips & Tricks"
format:
  html:
    css: custom.css
tbl-colwidths: [20,20,50,10]
---

This page collects miscellaneous tips and tricks for **tinyplot**. By
definition, these are workarounds---i.e., techniques that fall outside of
standard use cases or features that aren't (yet) natively supported by the
package. Please feel free to suggest or add more tips via our [GitHub repo](https://github.com/grantmcdermott/tinyplot/issues).

## Legend

### Opacity (`alpha`)

By default, the legend inherits the opacity of the plotted elements. In some
cases, this may be undesirable. For example, in the following plot, the legend
is too light:

```{r}
library(tinyplot)
x = rnorm(1e6)
y = x + rnorm(1e6)
z = sample(c("a", "b"), 1e6, replace = TRUE)
dat = data.frame(x, y, z)
tinyplot(y ~ x | z, data = dat, alpha = .1, pch = 19)
```

One solution is to draw our plot in two steps. First, we draw an empty plot with
the desired (zero transparency) legend. Second, we add the transparent points on
top of the existing canvas.

```{r}
tinyplot(y ~ x | z, data = dat, pch = 19, empty = TRUE)
tinyplot_add(empty = FALSE, alpha = .1)
```

## Labels

### Axis label rotation

When category labels are long or overlapping, users may want to rotate them for
readability. One option is fully perpendicular (90°) axis labels, which
`tinyplot` supports via themes, e.g. `tinytheme("clean", las = 2)`. If a user
wants finer control over the degree of rotation---say, 45°---this requires a bit
more manual effort since we do not support custom rotation out of the box. The
workaround involves three steps:

1. Suppress the default x-axis with `xaxt = "n"`.
2. Use `text()` to manually add rotated labels.
3. Optionally, clear the y-axis label by setting `ylab = ""`.

```{r}
library(tinyplot)
tinyplot(~cyl, data = mtcars, type = "barplot", xaxt = "n", ylab = "")
text(1:3, 0,
    labels = c("Four cylinders", "Six cylinders", "Eight cylinders"),
    srt = 45,          # rotate text 45 degrees
    adj = c(1.1, 1.5), # adjust text alignment
    xpd = TRUE)        # allow drawing outside plot region
```

Note that `adj` and `xpd` settings may require trial and error to position
labels correctly. Also, removing the x-axis label by setting `ylab = ""` is
unintuitive but currently necessary when using formulas like `~ cyl`.

P.S. Another option for long axis labels is to wrap them at a designated
character length. See the final example in the
[`tinylabel`](https://grantmcdermott.com/tinyplot/man/tinylabel.html)
documentation for an example.
