---
title: "Plot types"
---

`tinyplot` is a lightweight extension of R's base plotting system, designed to simplify and enhance data visualization. One of its key features is the `type` argument, which allows you to specify different types of plots easily. This tutorial will guide you through the various plot types available in `tinyplot`, demonstrating how to use the `type` argument to create a wide range of visualizations.

We will consider three categories of plot types:

1. Base types
2. `tinyplot` types
3. Custom types

We will use the built-in `airquality` dataset for our examples:

```{r}
library(tinyplot)

aq = transform(
  airquality,
  Month = factor(Month, labels = month.abb[unique(Month)]),
  Hot = Temp > median(Temp)
)
```

# Base types

The `type` argument in `tinyplot` supports all the standard plot types from base R's `plot` function. These are specified using single-character strings.

The default plot type is `"p"`, which creates a scatter plot of points.

```{r}
plt(Temp ~ Day | Month, data = aq, type = "p", main = "Points")
```

Use `"l"` to connect data points with lines.

```{r}
plt(Temp ~ Day | Month, data = aq, type = "l", main = "Lines")
```

* `b` (Points and Lines): Combines points and lines in the plot.
* `c` (Empty Points Joined by Lines): Plots empty points connected by lines.
* `o` (Overplotted Points and Lines): Overlaps points and lines.
* `s` and `S` (Stair Steps): Creates a step plot.
* `h` (Histogram-like Vertical Lines): Plots vertical lines resembling a histogram.
* `n` (Empty Plot): Creates an empty plot frame without data.

# `tinyplot` types

Beyond the base types, `tinyplot` introduces additional plot types for more advanced visualizations. 

Models:

- `loess`, `type_loess()`: Local regression curve.
- `lm`, `type_lm()`: Linear regression line.
- `glm`, `type_glm()`: Generalized linear model fit.

Shapes:

- `segments`: Draws line segments between pairs of points.
- `polygon`: Draws filled polygons.
- `errorbar`: Adds error bars to points; requires ymin and ymax.
- `pointrange`: Combines points with error bars.
- `ribbon`: Creates a filled area between ymin and ymax.
- `area`: Plots the area under the curve from y = 0 to y = f(x).
- `rect`: Draws rectangles; requires `xmin`, `xmax`, `ymin`, and `ymax`.

Visualizations:

- `jitter`, `type_jitter()`: Jittered points.
- `boxplot`, `type_boxplot()`: Creates a box-and-whisker plot.
- `histogram`: Creates a histogram of a single variable.
- `density`: Plots the density estimate of a variable.

There are two main ways to use `tinyplot` types. First, we can specify the type directly using a string shortcut in the `type` argument, such as `type="boxplot"`. For example, to create a box plot, use `type = "boxplot"`. 

Second, many `tinyplot` types are implemented by a `type_*()` function which includes arguments that allow us to customize the plot further. For example, the `type_jitter()` function includes arguments to control the amount of jittering, and the `type_glm()` includes an argument to control the width of the confidence interval to display. To see what arguments are available for each type, simply consult the type-specific documentation:

```r
?type_jitter
?type_glm
```

For example, we can add noise to data points using the `"jitter"` type. This allows us to avoid overplotting and can be useful to visualize discrete variables. On the left, we use the string shortcut with default settings. On the right, we use the `type_jitter()` function to select a smaller amount of jittering:

```{r}
#| layout-ncol: 2
plt(Temp ~ Month, data = aq, main = "Default jittering", 
    type = "jitter")

plt(Temp ~ Month, data = aq, main = "Modest jittering",
    type = type_jitter(amount = 0.05))
```

In this example, we use `type_glm()` to fit a logit regression model to the data, and display different confidence interval sizes:

```{r}
#| layout-ncol: 2
plt(Hot ~ Wind, data = aq, main = "Logit with 95% interval",
    type = type_glm(family = "binomial"))

plt(Hot ~ Wind, data = aq, main = "Logit with 50% interval",
    type = type_glm(family = "binomial", level = 0.5))
```

Most `tinyplot` types support grouping and faceting:

```{r}
plt(Temp ~ Wind | Month, data = aq, type = type_lm(se = FALSE))
```

# Custom types

It is easy to add custom types to `tinyplot`. To do this, we need to define 3 functions:

1. `data_*()`: Function factory.
    - Accepts a list of internal objects
        - Inputs must include `...`
        - `datapoints` Is the most important object. It is a data frame with the datapoints to plot. 
        - Other objects that can be modified by `data_*()` include: `by`, `facet`, `ylab`, `palette`
    - Returns a named list with modified versions of those objects.
2. `draw_*()`: Function factory.
    - Accepts information about data point values and aesthetics. 
        - Inputs must include `...`
        - The `i` prefix in argument names indicates that we are operating on a subgroup of the data, identified by `facet` or using the `|` operator in a formula.
        - Available arguments are: "ibg", "icol", "ilty", "ilwd", "ipch", "ix", "ixmax", "ixmin", "iy", "iymax", "iymin", "cex", "dots", "type", "x_by", "i", "facet_by", "split_data", "flip"
    - Returns a function which can call base R to draw the plot.
3. `type_*()`: A wrapper function that returns a named list with three elements:
    - `draw`
    - `data`
    - `name`

Here is a minimalist example of a custom type that logs both `x` and `y` and plots lines:

```{r}
type_log = function(base = exp(1)) {
  out = list(
    draw = draw_log(),
    data = data_log(base = base),
    name = "log"
  )
  class(out) = "tinyplot_type"
  return(out)
}

data_log = function(base) {
    fun = function(datapoints, ...) {
        datapoints$x = log(datapoints$x, base = base)
        datapoints$y = log(datapoints$y, base = base)
        datapoints = datapoints[order(datapoints$x), ]
        return(list(datapoints = datapoints, ...))
    }
    return(fun)
}

draw_log = function() {
    fun = function(ix, iy, icol, ...) {
        points(
            x = ix,
            y = iy,
            col = icol
        )
    }
    return(fun)
}

plt(mpg ~ wt | factor(am), data = mtcars,
    type = "p", main = "Raw")

plt(mpg ~ wt | factor(am), data = mtcars,
    type = type_log(), main = "Ln")

plt(mpg ~ wt | factor(am), data = mtcars,
    type = type_log(base = 10), main = "Log 10")
```

