---
title: "Plot types"
---

**tinyplot** is a lightweight extension of R's base plotting system, designed to
simplify and enhance data visualization. One of its key features is the `type`
argument, which allows you to specify different types of plots easily. This
tutorial will guide you through the various plot types available in
**tinyplot**, demonstrating how to use the `type` argument to create a wide
range of visualizations.

We will consider three categories of plot types:

1. Base types
2. **tinyplot** types
3. Custom types

We will use the built-in `airquality` dataset for our examples:

```{r}
library(tinyplot)

aq = transform(
  airquality,
  Month = factor(Month, labels = month.abb[unique(Month)]),
  Hot = Temp > median(Temp)
)
```

## Base types

The `type` argument in `tinyplot` supports all the standard plot types from
base R's `plot` function. These are specified using single-character strings.

- `"p"` (Points): Produces a scatter plot of points. This is (usually) the default plot type.
- `"l"` (Lines): Produces a line plot.
- `"b"` (Points and Lines): Combines points and lines in same the plot.
- `"c"` (Empty Points Joined by Lines): Plots empty points connected by lines.
- `"o"` (Overplotted Points and Lines): Overlaps points and lines.
- `"s"` and `"S"` (Stair Steps): Creates a step plot.
- `"h"` (Histogram-like Vertical Lines): Plots vertical lines resembling a histogram.
- `"n"` (Empty Plot): Creates an empty plot frame without data.

For example, we can use `"b"` to create a plot with combined points and lines.

```{r}
tinyplot(Temp ~ Day | Month, data = aq, type = "b", main = "Points and lines")
```

## tinyplot types

Beyond the base types, **tinyplot** introduces additional plot types for more
advanced visualizations. Each of these additional types are available either as
a convenience string (with default behaviour) or a companion `type_*()` function
(with options for customized behaviour).

Models:

- `"loess"` / `type_loess()`: Local regression curve.
- `"lm"` / `type_lm()`: Linear regression line.
- `"glm"` / `type_glm()`: Generalized linear model fit.
- `"spline"` / `type_spline()`: Cubic (or Hermite) spline interpolation.

Shapes:

- `area` / `type_area()`: Plots the area under the curve from `y` = 0 to `y` = f(`x`).
- `rect` / `type_area()`: Draws rectangles; requires `xmin`, `xmax`, `ymin`, and `ymax`.
- `ribbon` / `type_ribbon()`: Creates a filled area between `ymin` and `ymax`.
- `"segments"` / `type_segements()`: Draws line segments between pairs of points.
- `"polygon"` / `type_polygon()`: Draws polygons.
- `"polypath"` / `type_polypath()`: Draws a path whose vertices are given in `x` and `y`.
- `"errorbar"` / `type_errorbar()`: Adds error bars to points; requires `ymin` and `ymax`.
- `pointrange` / `type_pointrange`: Combines points with error bars.

Visualizations:

- `"boxplot"` / `type_boxplot()`: Creates a box-and-whisker plot.
- `"density"`: Plots the density estimate of a variable.
- `"histogram"` / `type_histogram()`: Creates a histogram of a single variable.
- `"jitter"` / `type_jitter()`: Jittered points.
- `"spineplot"` / `type_spineplot()`: Creates a spineplot or spinogram.

To see the difference between the convenience strings and their respective
companion `type_*` functions, let's quickly walk through two examples.

**Example 1: jittering.** We can add noise to data points using jittering. This
allows us to avoid overplotting and can be useful to visualize discrete
variables. On the left, we use the `"jitter` string shortcut with default
settings. On the right, we use the `type_jitter()` function to reduce the amount
of jittering.

```{r}
#| layout-ncol: 2
tinyplot(
  Temp ~ Month, data = aq, main = "Default jittering", 
  type = "jitter"
)

tinyplot(
  Temp ~ Month, data = aq, main = "Modest jittering",
  type = type_jitter(amount = 0.05)
)
```

**Example 2: Logit fit.** In this example, we use `type_glm()` to fit a logistic
regression model to the data, but with different confidence intervals.^[Note: If
we simply specified the `"glm"` convenience string on its own, we'd get a linear
fit since the default family is `gaussian`.]

```{r}
#| layout-ncol: 2
tinyplot(
  Hot ~ Wind, data = aq, main = "Logit with 95% interval",
  type = type_glm(family = "binomial")
)

tinyplot(
  Hot ~ Wind, data = aq, main = "Logit with 50% interval",
  type = type_glm(family = "binomial", level = 0.5)
)
```

To see what arguments are available for each type, simply consult the
type-specific documentation.

```r
?type_jitter
?type_glm
# etc
```

Finally, please note that all **tinyplot** types support grouping and faceting.

```{r}
tinyplot(Temp ~ Wind | Month, data = aq, facet = "by", type = "lm")
```

## Custom types

It is easy to add custom types to **tinyplot**. Users who need highly customized
plots, or developers who want to add support to their package or functions, only
need to define three simple functions: `data_typename()`, `draw_typename()`, and
`type_typename()`.

In this section, we explain the role of each of these functions and present a
minimalist example of a custom type. Interested readers may refer to
[the **tinyplot** source code](https://github.com/grantmcdermott/tinyplot)
to see many more examples, since each **tinyplot** type is itself implemented as a
custom type.

The three functions that we need to define for a new type are:

1. `data_*()`: Function factory.
    - Accepts a list of internal objects
        - Inputs must include `...`
        - `datapoints` Is the most important object. It is a data frame with the datapoints to plot. 
        - Other objects that can be modified by `data_*()` include: `by`, `facet`, `ylab`, `palette`
    - Returns a named list with modified versions of those objects.
2. `draw_*()`: Function factory.
    - Accepts information about data point values and aesthetics. 
        - Inputs must include `...`
        - The `i` prefix in argument names indicates that we are operating on a subgroup of the data, identified by `facet` or using the `|` operator in a formula.
        - Available arguments are: `ibg`, `icol`, `ilty`, `ilwd`, `ipch`, `ix`, `ixmax`, `ixmin`, `iy`, `iymax`, `iymin`, `cex`, `dots`, `type`, `x_by`, `i`, `facet_by`, `by_data`, `facet_data`, `flip`
    - Returns a function which can call base R to draw the plot.
3. `type_*()`: A wrapper function that returns a named list with three elements:
    - `draw`
    - `data`
    - `name`

Here is a minimalist example of a custom type that logs both `x` and `y` and
plots lines.

```{r}
#| layout-ncol: 2
type_log = function(base = exp(1)) {

  data_log = function() {
    fun = function(datapoints, ...) {
      datapoints$x = log(datapoints$x, base = base)
      datapoints$y = log(datapoints$y, base = base)
      datapoints = datapoints[order(datapoints$x), ]
      return(list(datapoints = datapoints, ...))
    }
    return(fun)
  }

  draw_log = function() {
    fun = function(ix, iy, icol, ...) {
      points(
        x = ix,
        y = iy,
        col = icol
      )
    }
    return(fun)
  }

  out = list(
    draw = draw_log(),
    data = data_log(),
    name = "log"
  )
  class(out) = "tinyplot_type"
  return(out)
}

tinyplot(mpg ~ wt | factor(am), data = mtcars,
  type = type_log(), main = "Ln")

tinyplot(mpg ~ wt | factor(am), data = mtcars,
  type = type_log(base = 10), main = "Log 10")
```

To underscore what we said above, the **tinyplot**
[source code](https://github.com/grantmcdermott/tinyplot/tree/main/R)
contains many examples of type constructor functions that should provide a
helpful starting point for custom plot types. Failing that, the **tinyplot**
team are always happy to help guide users on how to create their own types and
re-purpose existing **tinyplot** code. Just let us know by
[raising an issue](https://github.com/grantmcdermott/tinyplot/issues)
on our GitHub repo.
